Crime in Los Angeles: Homicide Increases in the COVID-19 Era

### 2025-05-01 1_DraftVersions/4_FourthDraft/Draft4.qmd
Reorganized repo

```{r,  warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}

query <- dbSendQuery(
  con,
  "
  SELECT
    YearWeek,
    MIN(Date) AS start_of_week,
    MAX(Date) AS end_of_week,
    COUNT(*)  AS Crimes
  FROM crime_la
  GROUP BY YearWeek
  ORDER BY YearWeek
  "
)
weekly_counts <- dbFetch(query, n = -1)
dbClearResult(query)

weekly_counts$start_of_week <- as.Date(weekly_counts$start_of_week)
weekly_counts$end_of_week   <- as.Date(weekly_counts$end_of_week)

weekly_counts$days_in_week <-
  as.numeric(difftime(weekly_counts$end_of_week,
                      weekly_counts$start_of_week,
                      units = "days")) + 1

weekly_counts <- subset(weekly_counts, days_in_week == 7)

weekly_counts$CrimesScaled <-
  weekly_counts$Crimes * (7 / weekly_counts$days_in_week)

weekly_counts$Index <- seq_len(nrow(weekly_counts))

march_2020_index <- which(weekly_counts$YearWeek == "2020-09")

plot(
  weekly_counts$Index,
  weekly_counts$CrimesScaled,
  type  = "l",
  col   = "blue",
  lwd   = 2,
  xlab  = "Weekly Index",
  ylab  = "Number of Crimes (Scaled to 7 Days)"
)

abline(
  h   = pretty(weekly_counts$CrimesScaled),
  v   = pretty(weekly_counts$Index),
  col = "gray",
  lty = "dotted"
)

abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)

text(
  x     = weekly_counts$Index[march_2020_index] - 80,
  y     = 10,
  labels = "March 2020",
  pos   = 3,
  col   = "red"
)

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget, Crm.Cd.Desc)

@@

write.csv(correlations, "corr.csv")

@@

write.csv(homicides_lat_long_yr, "homicides.csv")

>>>>

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget)

@@

@@

### 2025-05-01 1_DraftVersions/4_FourthDraft/Draft4_2.qmd
Reorganized repo

During the COVID-19 pandemic in 2021, homicide rates in Los Angeles increased substantially. A report by the Legal Defense Fund's Thurgood Marshall Institute found that this increase was associated with both pre-pandemic and pandemic-induced economic instability and inequalities.[@moore2022]

@@

To group this data geographically, we will organize the crimes by their latitude and longitude into the appropriate Census Blocks and ZIP Code Tabulation Areas (ZCTAs). Additionally, we will examine the impact of latitude and longitude themselves, as humans do not always behave by administrative unit boundaries and latitude and longitude by themselves might suggest a different pattern.[@krieger2002]

@@

To maximize the amount of data available for analysis, we used LA County socioeconomic data from 2019-

@@

We will also investigate the relationship between homicide and enforcement factors. In particular, we will look at the LAPD budget in comparison to the homicide rate to determine if increased police funding is effective in reducing the number of homicides.
```{r, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE)
```

@@

Figure 2 analyzes the number of crimes over time using time series. The time series shows stable weekly crime counts with occasional large spikes, likely due to shorter weeks. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024. The red line, which denotes March 2020, shows a decline in crimes, likely during the shutdown and quarantine, but it returned to normal levels quickly.

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget, Crm.Cd.Desc)

@@

Figure 9 shows the correlations between different variables that could affect the number of crimes committed. We see that the most significant correlations for the number of crimes are the unemployment rate, year, and number of evictions. Years is positively correlated with number of crimes, meaning that as time passed, the number of crimes increased. Similarly, as the number of evictions increases, so does the number of crimes. Interestingly, as the unemployment rate increases, the number of crimes decreases, which is the opposite effect of what was expected (one would anticipate that higher unemployment would result in more crime). The LAPD budget interestingly had a medium-strength positive correlation with number of crimes, suggesting we should further examine the increase in the LAPD budget compared to the increase in crimes.

@@

# Statistical Analyses

To examine the relationship between the number of homicides and various socioeconomic and enforcement factors, we will use a variety of statistical analyses to see if certain factors are more influential in causing increases or decreases in homicide rates.

@@

@@

# Appendices {.appendix}

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

```{r, output=FALSE,  message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

```{r  fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

```{r,  warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)

weekly_counts$Index <- seq_len(nrow(weekly_counts))
# Find the Index for March 2020
march_2020_index <- which(weekly_counts$YearWeek == "2020-09")  # Adjust '202009' to match your data's YearWeek format for March 2020

# Plot
plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")

# Add vertical line for March 2020
abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)

# Add a label for March 2020
text(
  x = weekly_counts$Index[march_2020_index] - 80, 
  y = 10, 
  labels = "March 2020", 
  pos = 3, 
  col = "red"
)
```

# since we do not have pre- and post- pandemic city of LA data, we are instead using LA county data for evictions. 
query <- dbSendQuery(con, "SELECT * FROM evictions_county_la")
LA_county_evictions <- dbFetch(query,n=-1)
dbClearResult(query)

# covid data
query <- dbSendQuery(con, "
  SELECT 
    CAST(strftime('%Y', ep_date) AS INTEGER) AS year,
    CAST(strftime('%m', ep_date) AS INTEGER) AS month,
   CAST(strftime('%d', ep_date) AS INTEGER) AS date, 
   cases_14day
  FROM covid_cases
")
covid_14day <- dbFetch(query,n=-1)
dbClearResult(query)
# since there is a 7 day lag in reporting, let's do monthly = cases on the 21st and on the 5th of the next month.
covid_14day$lagged <- covid_14day$date + 7

# next, unemployment rate
query <- dbSendQuery(con, "
  SELECT 
    u.month, 
    u.Year, 
    u.Value, 
    l.police_budget
  FROM unemployment_la AS u
  INNER JOIN lapd_budget AS l ON u.Year = l.year
")
unemployment_lapd_budget <- dbFetch(query,n=-1)
dbClearResult(query)

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, Value, num_crimes, police_budget) %>% rename("unemployment_rate" = "Value")

correlations_df3 <- correlations_df2 %>% inner_join(LA_county_evictions, join_by("year" == "Year", "month" == "Month")) %>% rename("num_evictions" = "Count")

correlations <- correlations_df3 %>% dplyr::select(year, month, unemployment_rate, num_crimes, cases, num_evictions, police_budget)

>>>>

During the COVID-19 pandemic in 2021, homicide rates in Los Angeles increased substantially. A report by the Legal Defense Fund's Thurgood Marshall Institute found that this increase was associated with both pre-pandemic and pandemic-induced economic instability and inequalities[@moore2022].

@@

To group this data geographically, we will organize the crimes by their latitude and longitude into the appropriate Census Blocks and ZIP Code Tabulation Areas (ZCTAs). Additionally, we will examine the impact of latitude and longitude themselves, as humans do not always behave by administrative unit boundaries and latitude and longitude by themselves might suggest a different pattern[@krieger2002].

@@

To maximize the amount of data available for analysis, we used LA County socioeconomic data for any correlation analysis and focused on years 2020-2023.

@@

We will also investigate the relationship between homicide and enforcement factors. In particular, we will look at the LAPD budget in comparison to the homicide rate to determine if increased police funding is effective in reducing the number of homicides.

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE)
```

@@

Figure 2 analyzes the number of crimes over time using time series. The time series shows stable weekly crime counts with occasional large spikes. One large increase occurs around Weekly Index 300, which is around 2016-17. This spike was attributed to a resurgence in gang violence and the growth in homelessness. The red line, which denotes March 2020, shows a decline in crimes, likely during the shutdown and quarantine, but it returned to normal levels quickly. Since then, crimes have decreased recently, but this could be caused by some issues with 2024 data, as the LAPD said they changed their reporting standards: "Records Management System for reporting crimes and arrests" and that they were experiencing issues in 2024.

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget, num_crimes)

@@

Figure 9 shows the correlations between different variables that could affect the number of crimes committed. We see that the most significant correlations for the number of crimes are the unemployment rate, year, and number of evictions. Years is positively correlated with number of crimes, meaning that as time passed, the number of crimes increased. Similarly, as the number of evictions increases, so does the number of crimes. Interestingly, as the unemployment rate increases, the number of crimes decreases, which is the opposite effect of what was expected (one would anticipate that higher unemployment would result in more crime). Studies have shown that crime often increases during periods of low unemployment and that employed people commit many crimes[@tarling1982]. The LAPD budget interestingly had a weak-strength positive correlation with number of crimes.

@@

# Modeling

To examine the relationship between the number of homicides and various socioeconomic and enforcement factors, we will use a variety of models to see if certain factors are more influential in causing increases or decreases in homicide rates.

@@

# Killer Plot Visualization

The killer plot visualizes crime spatially and over time, stylized based on the Hollywood Walk of Fame since that’s an iconic part of Los Angeles. The plot shows the average homicides by latitude and longitude, and you can select a different year. You can see that the rates have remained higher in the southeastern areas than the rest of the city, similar to the density map earlier. Still, the rates in the northwestern areas are higher in 2023, relative to the rest of the city, compared to previous years where there was a clear difference.

![Understand crimes spatially through the killer plot.](kp1.png)

You can also view different variables over time. For example, you can select a range of years to view a particular crime over time, such as homicide. You can see that homicide rates have been higher in the last few years. You can also look at the decrease in COVID cases since the pandemic, the variability in the number of evictions over time, the increase in the LAPD budget, and the decrease in the unemployment rate since the pandemic.

![Understand crimes and other variables over time in a timeline-style view.](kp2.png)

# Conclusion

Through data visualization and modeling, we have discerned some interesting spatiotemporal patterns in crimes over the last 15 years. Figure 2, for example, showed that there have been two spikes in crimes since 2010, one in 2016-17 and one a few weeks after March 2020. The increase in 2016-17 is explained by difficulties managing gang activities and a growing homeless population in Los Angeles[@chang2016]. The second increase can likely be attributed to the pandemic factors described in the Thurgood Marshall Institute report.

Correlations found that evictions were strongly correlated with the number of crimes, suggesting that as the number of evictions increased, crimes also increased. This aligns with the finding of the report about homicides, that "social and economic instability and heightened inequalities generated by the COVID-19 pandemic" explains the increase in homicides. While evictions in Los Angeles in mid-2020 were lower due to renter protections, by November 2020 the levels once again increased, and reached pre-pandemic levels by 2022[@covid-19]. However, there was no evidence that COVID-19 case levels had a strong relationship, and unemployment rate had a strong negative relationship with crime rate, suggesting that as unemployment decreases, the number of crimes increases. This, while surprising, aligns with past findings in the literature that employed people commit more crimes.

The Thurgood Marshall Institute report also noted that increased police funding was not effective in reducing homicide rates. The increase in LAPD funding over the years was a moderately strong positive relationship between time and funding, while crime had a strong positive relationship over time. Additionally, a linear model found year to be a significant predictor of homicides, suggesting that over time, the number of homicides has increased. This, combined with the weak correlation between crime and LAPD budget, supports this assertion and suggests that increasing police funding might not be an effective strategy.

Finally, we found evidence of spatial trends in crime and homicide rates. Table 4 shows that, even when controlling for year, latitude and longitude are significant predictors of homicide rate. Namely, homicides are more common in the south and eastern parts of Los Angeles.

In conclusion, we found that crimes have increased through the pandemic years, like the findings in the Thurgood Marshall Institute report. We found that in contrast to our initial hypothesis, socioeconomic conditions have not necessarily improved in LA. Namely, evictions have returned and surpassed pre-pandemic levels and income inequality appears to be worse than before as more of LA becomes wealthier but certain neighborhoods remain below the poverty line. Higher homicide counts occur in areas with lower income, though we've found evidence that this pattern is changing as homicides have decreased in lower income areas and increased in higher income areas in the past few years. Ultimately, in order to reduce homicide and crime levels, the City of Los Angeles must do more to combat socioeconomic factors that are associated with increases in crimes, like housing insecurity.

@@

### 2025-05-01 2_Slides/SlidesFinal.qmd
Reorganized repo

---
title: "Crime in Los Angeles: Homicide Increases in the COVID-19 Era"
subtitle: "Bianca Schutz, Kevin Hu, Kaitlyn Kirt, Marco Stine"
format:
  revealjs:
# embed-resources: true
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    css: styles.css
    footer: Group 5 - STAT 405 - Bianca Schutz, Kevin Hu, Kaitlyn Kirt, Marco Stine - Crime in Los Angeles
server: shiny
---

@@

   grid.newpage()
   grid.rect(gp = gpar(fill = "gray15", col = NA))  # No border, black fill
   for (star in 1:nrow(df)) {
        row <- as.list(df[star,])
        grid.draw(create_star(scale_factor = row$scale, center = c(row$LONstd, row$LATstd))) 
        grid.text(label = paste0(row$LAT, ", ", row$LON),
    x = unit(row$LONstd, "npc"),
    y = unit(row$LATstd + 0.05, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 10))
        }
   grid.rect(x = 0.5, y = 0, width = 1, height = 0.125, just = "bottom", gp = gpar(col = "white", fill = "gray15"))

        grid.text(label = "Average Homicides",
    x = unit(.125, "npc"),
    y = unit(0.0625, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 14))

      grid.draw(create_star(scale_factor = 1, center = c(.25, .0625)))

      grid.draw(create_star(scale_factor = 2, center = c(.375, .0625)))

      grid.draw(create_star(scale_factor = 3, center = c(.6, .0625)))

      grid.draw(create_star(scale_factor = 4, center = c(.85, .0625)))

      grid.text(label = as.character(labels[1]), x = unit(.275, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[2]), x = unit(.475, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[3]), x = unit(.725, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[4]), x = unit(.925, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11)) 
} else {
      df <- selectedKillerData()
      print(head(df))
      rows <- nrow(df)
      print(rows)
      col_row <- if (rows <= 13) {
        c(rows, 1)
      } else if (rows == 14) {
        c(7, 2)
      } else if (rows == 15) {
        c(5, 3)
      } else if (rows == 16) {
        c(4, 4)
      } else {
        c(ceiling(sqrt(rows)), ceiling(rows / ceiling(sqrt(rows))))
      }

      stars <- generate_star_centers(col_row[1], col_row[2])
      print(nrow(stars))
      labels <- levels(standardize_scale(df$variable))
      print(labels)
      df_sorted <- df[order(df$year), ]

      stars$scale <- df_sorted$bins
      stars$year <- df_sorted$year
      print(stars)

      grid.newpage()
      grid.rect(gp = gpar(fill = "gray15", col = NA))  # No border, black fill

      for (star in 1:nrow(stars)) {
        row <- as.list(stars[star,])
        grid.draw(create_star(scale_factor = row$scale, center = c(row$x, row$y)))

        grid.text(label = as.character(row$year),
    x = unit(row$x, "npc"),
    y = unit(row$y - 0.05, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 10))
      }
        grid.rect(x = 0.5, y = 0, width = 1, height = 0.125, just = "bottom", gp = gpar(col = "white", fill = "gray15"))

        grid.text(label = paste(stringr::str_to_title(input$dataset), "Level"),
    x = unit(.1, "npc"),
    y = unit(0.0625, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 14))

      grid.draw(create_star(scale_factor = 1, center = c(.225, .0625)))

      grid.draw(create_star(scale_factor = 2, center = c(.375, .0625)))

      grid.draw(create_star(scale_factor = 3, center = c(.6, .0625)))

      grid.draw(create_star(scale_factor = 4, center = c(.85, .0625)))

      grid.text(label = as.character(labels[2]), x = unit(.275, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[3]), x = unit(.475, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[4]), x = unit(.725, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[5]), x = unit(.925, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11)) }
    })
```

>>>>

---
title: "Crime in Los Angeles: Homicide Increases in the COVID-19 Era"
format:
  revealjs:
    theme: solarized
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    footer: Group 5 - STAT 405 - Bianca Schutz, Kevin Hu, Kaitlyn Kirt, Marco Stine - Crime in Los Angeles
    title-slide-attributes:
      data-background: "#0f172a"
      data-transition: fade
server: shiny
---

@@

   draw_spatial_killer(df, labels)
} else {
      df <- selectedKillerData()
      rows <- nrow(df)
      col_row <- if (rows <= 13) {
        c(rows, 1)
      } else if (rows == 14) {
        c(7, 2)
      } else if (rows == 15) {
        c(5, 3)
      } else if (rows == 16) {
        c(4, 4)
      } else {
        c(ceiling(sqrt(rows)), ceiling(rows / ceiling(sqrt(rows))))
      }

      stars <- generate_star_centers(col_row[1], col_row[2])
      labels <- levels(standardize_scale(df$variable))
      df_sorted <- df[order(df$year), ]

      stars$scale <- df_sorted$bins
      stars$year <- df_sorted$year

      draw_timeline_killer(stars, labels, input$dataset)
       }
    })
```
